---
title: "Viewing and Modifying Priors}"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
    theme: united
vignette: >
  %\VignetteIndexEntry{Viewing and Modifying Priors}}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r}
#detach("package:malclimsim", unload = TRUE)
#devtools::install_github("https://github.com/SwissTPH/malclimsim")
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
library(malclimsim)    # your package
library(mcstate)       # for pmcmc_parameter() and pmcmc_parameters$new()
library(ggplot2)       # for plotting
```

# Overview

This vignette demonstrates how to use the following functions for managing priors in **malclimsim**:

1.  `return_default_priors()`\
2.  `build_priors()`\
3.  `view_priors()`\
4.  `plot_priors()`

We will:

-   Extract the raw defaults from `return_default_priors()`.\
-   Use `build_priors()` to construct `mcstate::pmcmc_parameter()` objects (either “pure defaults” or with overrides).\
-   Inspect those priors with `view_priors()`.\
-   Plot them with `plot_priors()`.\
-   Override one or more priors by passing an `override_priors` list into `build_priors()`.

> **Note on terminology**\
> - `return_default_priors()` returns a *raw* R list of lists (each element has `initial`, `min`, `max`, and `prior`).\
> - `build_priors()` takes that raw list plus any `override_priors` and returns a named list of `pmcmc_parameter` objects.\
> - `view_priors()` and `plot_priors()` each accept a `priors` argument that should be a list of `pmcmc_parameter` objects (built by `build_priors()`).

------------------------------------------------------------------------

# 1. Building a Minimal Example

We first create a completely self‐contained example:

1.  Pull all raw defaults via `return_default_priors()`.\
2.  Build `param_inputs` (a named list of initial values).\
3.  Create a dummy identity `proposal_matrix` whose rownames exactly match the parameter names.\
4.  Choose a small subset of parameters to “estimate” (`params_to_estimate`).

```{r minimal-setup}
# 1. Get the raw default prior specifications:
default_priors <- return_default_priors()

# 2. Build `param_inputs`: a named list where each element is the raw “initial” for that parameter
param_inputs <- lapply(default_priors, function(x) x$initial)

# 3. Build a dummy proposal_matrix:
#    - identity matrix of size = number of parameters
#    - rownames (and colnames) exactly match the names of default_priors
all_param_names <- names(default_priors)
proposal_matrix <- diag(length(all_param_names))
rownames(proposal_matrix) <- colnames(proposal_matrix) <- all_param_names

# 4. Pick a small subset of parameters to demonstrate:
params_to_estimate <- c("phi", "a_R", "qR", "z", "eff_SMC", "size")

# Show what we have in param_inputs and proposal_matrix:
str(param_inputs[params_to_estimate], max.level = 1)
str(proposal_matrix[params_to_estimate, params_to_estimate])
```

# 2. Building “Pure Default” Priors

Next, we build a list of `pmcmc_parameter` objects from the package defaults, using `build_priors()`. Since we do *not* supply any `override_priors`, `build_priors()` will:

1.  Internally call `return_default_priors()` to get the defaults.\
2.  Wrap each element (for the names in `params_to_estimate`) in `mcstate::pmcmc_parameter()`.

```{r build-default-priors}
# Build a list of pmcmc_parameter() objects (pure defaults):
default_mcmc_priors <- build_priors(
  param_inputs,
  proposal_matrix,
  params_to_estimate,
  override_priors = NULL
)

# Show the names and one example element:
names(default_mcmc_priors)
default_mcmc_priors$phi
```

# 3. Viewing Default Priors

We can now call `view_priors()` to get a **data.frame** summarizing each prior’s:

-   **Name**\
-   **Initial** value\
-   **Min** / **Max** bounds\
-   A textual **Description** of the prior function (as a string)

Because `view_priors()` expects a list of `pmcmc_parameter` objects, we pass it `default_mcmc_priors` via the `priors` argument:

```{r view-default-priors}
priors_df <- view_priors(
  param_inputs       = param_inputs,
  proposal_matrix    = proposal_matrix,
  params_to_estimate = params_to_estimate,
  priors             = default_mcmc_priors
)
priors_df
```

# 4. Plotting Default Priors

Next, we visualize each prior over its support.

```{r plot-default-priors, fig.width=7, fig.height=5}
prior_plot <- plot_priors(
  param_inputs       = param_inputs,
  proposal_matrix    = proposal_matrix,
  params_to_estimate = params_to_estimate,
  priors             = default_mcmc_priors
)
print(prior_plot)
```


# 5. Modifying One or More Priors

Suppose we want to change the default for **two** parameters:

1.  **phi**:
    -   Originally: Uniform(0.01, 2).\
    -   We want to switch to Beta(40, 12) on [0.01, 1].
2.  **qR**:
    -   Originally: Normal(mean = 0.24, sd = 0.5), unbounded except by `(min = 1e-6, max = 0.5)`.\
    -   We want to force `qR ∈ [0.01, 0.5]` with Uniform(0.01, 0.5).

To do that, we build an `override_priors` list in the same structure as `return_default_priors()`. Then we call `build_priors()` again—but passing our `override_priors`. Anything we specify in `override_priors` will replace the package default for that parameter; everything else falls back to the original.

```{r override-priors}
override_priors <- list(
  phi = list(
    initial = 0.5,
    min     = 0.01,
    max     = 1.0,
    prior   = function(p) dbeta(p, 40, 12, log = TRUE)
  ),
  qR  = list(
    initial = 0.1,
    min     = 0.01,
    max     = 0.5,
    prior   = function(p) dunif(p, 0.01, 0.5, log = TRUE)
  )
)

updated_mcmc_priors <- build_priors(
  param_inputs,
  proposal_matrix,
  params_to_estimate,
  override_priors = override_priors
)

updated_priors_df <- view_priors(
  param_inputs       = param_inputs,
  proposal_matrix    = proposal_matrix,
  params_to_estimate = params_to_estimate,
  priors             = updated_mcmc_priors
)
updated_priors_df
```


Notice how:

-   **phi** now has `Initial = 0.5`, `Min = 0.01`, `Max = 1.0`, and a Beta(40,12).\
-   **qR** now has `Initial = 0.1`, `Min = 0.01`, `Max = 0.5`, and a Uniform(0.01,0.5).\
-   All other four parameters remain unchanged from the original defaults.

# 6. Re‐Plotting the Updated Priors

Finally, let’s re‐plot our six priors so you can see how **phi** and **qR** have changed shape:

```{r plot-updated-priors, fig.width=7, fig.height=5}
updated_plot <- plot_priors(
  param_inputs       = param_inputs,
  proposal_matrix    = proposal_matrix,
  params_to_estimate = params_to_estimate,
  priors             = updated_mcmc_priors
)
print(updated_plot)
```

If you want to integrate directly into the main inference pipeline (`inf_run()`), you can pass `override_priors` as an argument there:

```{r eval=FALSE}
res <- inf_run(
  model              = my_model,
  param_inputs       = param_inputs,
  control_params     = control_params,
  params_to_estimate = params_to_estimate,
  proposal_matrix    = proposal_matrix,
  adaptive_params    = adaptive_params,
  start_values       = start_values,
  dates              = c("2020-01-01", "2022-12-31"),
  obs_config         = obs_config,
  override_priors    = override_priors
)
```
